---
title: "CT Pizza Map"
author: "Evan Killion"
date: "2026-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

First, going to do some set up, including:
- Cleaning environment
- Setting up packages

```{r project setup}
## Remove any lists in environment
rm(list = ls())

## Packages
library(dplyr)
library(tidyr)
library(sf)
library(sfnetworks)
library(dodgr)
library(stringr)

## Directories
project_dir = paste0(getwd(), "/")
output_dir = paste0(getwd(), "/Outputs/")
```

## Data

Now, I need to read in data. At first, I'll only test the Apizza joints (less combinations), but I'll read everything in for good measure:
- Road Network
- Census blocks
- Table of block-apizza restaurant correspondences
- All pizza joints (not right now)
- Apizza pizza joints
- Greek pizza joints (not right now)
- Neapolitan pizza joints (not right now)
- Tavern pizza joints (not right now)
- Thick pizza joints (not right now)
- 


```{r read data}

## Read in road network
roads = st_read(paste0(project_dir, "CT Roads with Speed Limits Exploded.gpkg"))

## Read in Pizza joints
# all_pizza = st_read(paste0(project_dir, "Pizza Spots Fixed Locations.gpkg"))
apizza = st_read(paste0(project_dir, "Apizza Pizza Shops.gpkg"))
# greek = st_read(paste0(project_dir, "Greek Pizza Shops.gpkg"))
# neapolitan = st_read(paste0(project_dir, "Neapolitan Pizza Shops.gpkg"))
# tavern = st_read(paste0(project_dir, "Tavern Pizza Shops.gpkg"))
# thick = st_read(paste0(project_dir, "Thick Pizza Shops.gpkg"))

## Read in census blocks
census_blocks = st_read(paste0(project_dir, "Clean Census Blocks.gpkg"))

## Read in correspondence table
blocks_apizza = read.csv(paste0(project_dir, "Blocks with Apizza Joints.csv"), header = TRUE, check.names=FALSE)

## Read in state boundary file
state_shape = st_read(paste0(project_dir, "/tl_2025_us_state/tl_2025_us_state.shp"))
```

## Shapefiles

Next, I'm going to work on prepping shapefiles for processing

```{r shapefiles}

# Pizza
apizza = apizza %>% 
  st_transform(4326) %>% 
  mutate(id = row_number()) %>% 
  select(unique_id, Name, geom) %>% 
  rename(pizza_id = unique_id)

# Census geog
census_blocks = census_blocks %>% 
  st_transform(4326) %>%
  st_centroid() %>% 
  select(GEOID20, geom)

# CT boundary
conn = state_shape %>% 
  filter(STUSPS == "CT")

```

## Correspondence table

After processing the shapefiles a bit further to get them ready for network analysis, I'm going to work on getting the correspondence table ready to be used as a filter for network analysis.

```{r correspondence}

# Remove extraneous variables
blocks_apizza = blocks_apizza %>% 
  select(GEOID20, Name, unique_id) %>% 
  rename(pizza_id = unique_id) %>% 
  mutate(GEOID20 = str_pad(GEOID20, width = 15, pad = "0"))

# Check that variables line up
## Variable type
class(blocks_apizza$GEOID20) == class(census_blocks$GEOID20)
class(blocks_apizza$pizza_id) == class(apizza$pizza_id)

## Spot check values
head(blocks_apizza$GEOID20) %in% census_blocks$GEOID20
head(blocks_apizza$pizza_id) %in% apizza$pizza_id
```

# Network prep

Here comes the fun part! Now that I've got all my files loaded in, I'll use dodgr to perform network analysis. I created a table of the census blocks with the five closest pizza joints in QGIS so I can limit the possibilities and I've made sure my road network has speed (and now a weighted distance based on that) to get closer to true travel times. 

```{r network}

# Road network
## CRS
roads = roads %>% 
  st_transform(4326)

## Find length
roads = roads %>% 
  ### Needs to be in projected CRS for length to work
  mutate(d = as.numeric(st_length(st_transform(., 6433)))) %>% 
  ### Transform back
  st_transform(4326) %>% 
  mutate(d_weighted = d/(speed*0.44704))

## Roads to network, then to edges and nodes
### Network
net = as_sfnetwork(roads, directed = FALSE)

### Edges
edges = net %>% 
  activate("edges") %>% 
  st_as_sf() %>% 
  mutate(
    from_id = as.character(from),
    to_id = as.character(to)
  ) %>% 
  st_drop_geometry() %>% 
  select(from_id, to_id, d, d_weighted)
  
### Nodes
nodes = net %>% 
  activate("nodes") %>% 
  st_as_sf() %>% 
  mutate(node_id = as.character(row_number())) %>% 
  bind_cols(st_coordinates(.)) %>%
  st_drop_geometry() %>% 
  select(node_id, X, Y) %>% 
  rename(lon = X, lat = Y)
  
### Add node coords to edges table 
edges = edges %>% 
  left_join(nodes, by = c("from_id" = "node_id")) %>% 
  rename(from_lon = lon, from_lat = lat) %>% 
  left_join(nodes, by = c("to_id" = "node_id")) %>% 
  rename(to_lon = lon, to_lat = lat) %>% 
  mutate(edge_id = row_number()) %>% 
  relocate(edge_id, .before = from_id)

### Filter out unconnected edges
edges_connected_test = dodgr_components(edges) 

edges_connected = edges_connected_test %>% 
  filter(component == 1)

edges_disconnect = edges_connected_test %>% 
  filter(component != 1)

disconnect_sf = edges_disconnect %>% 
  rowwise() %>%
  mutate(geometry = st_sfc(
    st_linestring(
      matrix(c(from_lon, from_lat, to_lon, to_lat), nrow = 2, byrow = TRUE)
    )
  )) %>%
  ungroup() %>%
  st_as_sf(crs = 4326)

st_write(disconnect_sf, paste0(project_dir, "edges_disconnected.shp"))

### Make sure bidirectional
#### Make an inverse version
reverse_edges <- edges_connected %>%
  rename(
    from_id = to_id,
    to_id = from_id,
    from_lon = to_lon,
    to_lon = from_lon,
    from_lat = to_lat,
    to_lat = from_lat
  )

#### Combine
edges_bidirectional <- bind_rows(edges_connected, reverse_edges) %>%
  distinct(from_id, to_id, .keep_all = TRUE)
  
```

# Network Analysis

Now that the graph is prepped, I can go into actually computing the distances between places from the table!

```{r network analysis}

# Define origins
## Put in order of correspondence table so that I have two ordered matrices of coords for pairwise distance matrix
origin_coords = blocks_apizza %>% 
  left_join(st_drop_geometry(census_blocks) %>% 
              bind_cols(st_coordinates(census_blocks)),
            by = "GEOID20") %>% 
  select(X, Y) %>% 
  as.matrix()
## Make colnames x and y instead of X and Y so dodgr doesn't get pissy
colnames(origin_coords) = c("x", "y")

# Define destinations
## Put in order of correspondence table so that I have two ordered matrices of coords for pairwise distance matrix
dest_coords = blocks_apizza %>% 
  left_join(st_drop_geometry(apizza) %>% 
              bind_cols(st_coordinates(apizza)),
            by = "pizza_id") %>% 
  select(X, Y) %>% 
  as.matrix()
## Make colnames x and y instead of X and Y so dodgr doesn't get pissy
colnames(dest_coords) = c("x", "y")

# Verify CRS
st_crs(origin_coords)
st_crs(dest_coords)
st_crs(roads)
st_crs(net)

# Verify bounding box
st_bbox(apizza)
st_bbox(roads)
st_bbox(census_blocks)

# Test run, first and last row of edges
test_from <- matrix(c(edges_bidirectional$from_lon[1], edges$from_lat[1]), nrow = 1)
colnames(test_from) = c("x", "y")
test_to <- matrix(c(edges_bidirectional$to_lon[100], edges$to_lat[100]), nrow = 1)
colnames(test_to) = c("x", "y")
## to is 41.420705, -72.491574, near gilette castle
## from is 41.341021, -73.082404, ansonia
## Test the distances
dodgr_distances(edges_bidirectional, from = test_from, to = test_to)

# Run network analysis matrix
dist_matrix = dodgr_distances(
  edges_bidirectional, 
  from = origin_coords,
  to = dest_coords,
  ## Running pairwise to only test combinations in correspondence table
  pairwise = TRUE
)

```

# Final pairs

Now that the actual network analysis is out of the way, I need to find the final pair for each block. I'll join the ordered result back onto the correspondence table, group by shortest distance (weighted by time), 

```{r final pairs}

# Join matrix to correspondence table
pairs = blocks_apizza %>% 
  mutate(travel_time_sec = dist_matrix) %>% 
  mutate(travel_time_min = travel_time_sec/60)

# Check some pairs
head(pairs, n = 25)
##1:
##2:
##3:
##4:
##5:


# Group to find final pairs
pairs = pairs %>% 
  group_by(GEOID20) %>% 
  slice_min(travel_time, n = 1) %>% 
  ungroup

# Check final output
slice_sample(pairs, n = 10)

```

# Visualization

With the final pairs produced, I'll want to create a leaflet map with:
- CT state boundary
- Blocks colored by closest restaurant (no outlines around blocks so as to make it look clean)
- Restaurant points

```{r leaflet maps}



```
